---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

The following code is needed to load the activity.csv dataset.

```{r loaddata, echo=TRUE}
activity <- read.csv("activity.csv")
activity$date <- as.Date(as.character(activity$date), "%Y-%m-%d")
dailystep <- aggregate(activity$steps, by=list(activity$date), FUN=sum)
intervalstep <- aggregate(activity$steps, by=list(activity$interval), FUN=mean, 
                          na.rm=TRUE)
names(dailystep) <- c("date", "steps")
names(intervalstep) <- c("interval", "steps")
```

## What is mean total number of steps taken per day?

The following shows the histogram of the average total number of steps taken 
per day.

```{r histogram, echo=TRUE}
hist(dailystep$steps, xlab = "No. of Steps", ylab = "Number of days",
     main = "Histogram of no. of steps taken per day", ylim = c(0, 30))
meanstep <- round(mean(dailystep$steps, na.rm = TRUE), digits=2)
medianstep <- round(median(dailystep$steps, na.rm = TRUE), digits=2)
```

The mean of total number of steps taken per day is 
**`r format(meanstep, scientific=FALSE, big.mark=',')`** 
and the median of total number of steps taken per day is 
**`r format(medianstep, scientific=FALSE, big.mark=',')`**.

## What is the average daily activity pattern?

The following diagram shows a line plot of the 5-minute interval (x-axis) and 
the average number of steps taken, averaged across all days (y-axis).

```{r lineplot, echo=TRUE}
library("ggplot2")
g <- ggplot(intervalstep, aes(interval, steps))
g + geom_line(color="blue") + labs(x = "Interval") + labs(y = "Number of steps")
maxstep <- subset(intervalstep, steps==max(intervalstep$steps))
```

The 5-minute interval **`r maxstep$interval`**, on average across all the days 
in the dataset, contains the maximum number of steps. The maximum number of step
is **`r format(round(maxstep$steps, digits=2), scientific=FALSE, big.mark=',')`**.

## Imputing missing values

There are **`r nrow(activity[is.na(activity$steps),])`** missing values in the 
dataset.  

I would like to use **the median for that 5-minute interval** for filling in all
the missing values in the dataset.  

To create the new dataset, I ran the following code. The first 20 lines of the 
result is shown at the second block.

```{r fillNAvalues, echo=TRUE}
missingactivity <- activity[is.na(activity$steps),]
nonmissingactivity <- activity[!is.na(activity$steps),]
medianintervalstep <- aggregate(activity$steps, by=list(activity$interval), 
                                FUN=median, na.rm=TRUE)
names(medianintervalstep) <- c("interval", "steps")
filledactivity <- merge(x = missingactivity, y = medianintervalstep, 
                        by = "interval", all.x = TRUE)
filledactivity$steps = filledactivity$steps.y
filledactivity <- filledactivity[,c("interval","date","steps")]
filledactivity <- rbind(filledactivity, nonmissingactivity)
head(filledactivity, 20)
```
There are **`r nrow(filledactivity[is.na(filledactivity$steps),])`** missing 
values in the dataset after filling missing values. 

The following shows the histogram of the average total number of steps taken 
per day (with missing value being filled).

```{r filledhistogram, echo=TRUE}
filleddailystep <- aggregate(filledactivity$steps, by=list(activity$date), 
                             FUN=sum)
names(filleddailystep) <- c("date", "steps")
hist(filleddailystep$steps, xlab = "Number of steps", ylab = "Number of days",
     main = "Histogram of no. of steps taken per day (without missing data)", 
     ylim = c(0, 30))
filledmeanstep <- round(mean(filleddailystep$steps, na.rm = TRUE), digits=2)
filledmedianstep <- round(median(filleddailystep$steps, na.rm = TRUE), digits=2)
```

After filling the missing value, 
the mean of total number of steps taken per day changes from 
**`r format(meanstep, scientific=FALSE, big.mark=',')`** to 
**`r format(filledmeanstep, scientific=FALSE, big.mark=',')`** 
and the median of total number of steps taken per day changes from
**`r format(medianstep, scientific=FALSE, big.mark=',')`** to
**`r format(filledmedianstep, scientific=FALSE, big.mark=',')`**.

## Are there differences in activity patterns between weekdays and weekends?

The new factor variable **week** is created, the following shows the first 6 
records of the dataset.

```{r weekday, echo=TRUE}
weekend <- weekdays(filledactivity$date, abbreviate=TRUE) %in% c("Sat","Sun")
filledactivity[weekend, "week"] <- "weekend"
filledactivity[!weekend, "week"] <- "weekday"
head(filledactivity)
```

The following shows the panel plot of the 5-minute interval and the average 
number of steps taken, averaged across all weekday days or weekend days.

```{r weekdayplot, echo=TRUE}
intervalweekstep <- with(filledactivity, aggregate(steps, 
                                                   by=list(interval, week), 
                                                   FUN=mean))
names(intervalweekstep) <- c("interval", "week", "steps")
g <- ggplot(intervalweekstep, aes(interval, steps))
g + geom_line(color="blue") + facet_wrap(~ week, nrow=2) + labs(x = "Interval") + 
        labs(y = "Number of steps")
```
